Automatically proving equivalence by type-safe reflection

### 2016-01-15 paper-JFP-DTP/content/1-introduction.tex
Added the source file for the example with Nat
paper : use of listings and corrections

For example, when working with natural numbers, one might need to prove the lemma $\forall\ xy:Nat,\ x*x\ +\ 3*x\ +\ x*y\ =\ x*(3+y+x)$, and when working with lists and concatenation of lists, one might need to prove something like $\forall\ l1\ l2\ l3:List A,\ ((l1\ ++\ (l2 ++ l3))\ ++\ l4\ =\ (l1++l2)\ ++\ (l3++l4)$. Proving this kind of auxiliary lemmas is time consuming, is much too specific for being reusable, and isn't the interesting part of formal certification. Therefore, we would like to be able to automatize a far as possible the constructions of proofs for this kind of lemmas for the Idris language, a dependently typed language which doesn't contain many proof automations at the moment.

A common situation in formal certification is the following. We have a function $f\ :\ A \rightarrow B$, and two functions $sem_A\ :\ A \rightarrow SEM$ and $sem_B\ : \ B \rightarrow SEM$ of interpretation which associate a semantic to each type A and B. If this semantic says everything we care about for the behaviour of $f$, then the lemma of correctness of f can be expressed as $forall\ a:A,\ semA\ a = semB (f\ a)$.
All the equality symbols $=$ we have encountered so far refer to the propositional equality. We remind the readers in the next subsection how this equality is defined in Idris, and more generally in Intentional Type Theories, and how it differs from the --more primitive-- judgemental equality.

@@

The judgemental equality has to be included in the propositional equal, because what is equal by definition should be provably equal.
This is accomplished by giving constructors to the type $Id(a,a)$ and nothing when "$a$ is not $b$".
\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
Id (A:Type) : A -> A -> Type :=
  Refl : forall a:A, Id a a
\end{verbatim}
\end{center}
\caption{The type Id}
\figrule
\end{figure}

@@

\subsection{A motivating little example}
We introduce a little example which shows how these proof obligations appear when we index a type by values like natural numbers.
Let's define the types of $Bit$ and $Binary$ number, both indexed over the value they represent, expressed in base ten as a natural number.
\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
data Bit : Nat -> Type where
     b0 : Bit Z
     b1 : Bit (S Z)

data Binary : (width : Nat) -> (value : Nat) -> Type where
     zero : Binary Z Z
     (#) : Binary w v -> Bit bit -> Binary (S w) (bit + 2 * v)
\end{verbatim}
\end{center}
\caption{Bit and binary number}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
addBit : Bit x -> Bit y -> Bit c -> (bX ** (bY ** 
              (Bit bX, Bit bY, c + x + y = bY + 2 * bX)))
addBit b0 b0 b0 = (_ ** (_ ** (b0, b0, _)))
addBit b0 b0 b1 = (_ ** (_ ** (b0, b1, _)))
addBit b0 b1 b0 = (_ ** (_ ** (b0, b1, _)))
addBit b0 b1 b1 = (_ ** (_ ** (b1, b0, _)))
addBit b1 b0 b0 = (_ ** (_ ** (b0, b1, _)))
addBit b1 b0 b1 = (_ ** (_ ** (b1, b0, _)))
addBit b1 b1 b0 = (_ ** (_ ** (b1, b0, _)))
addBit b1 b1 b1 = (_ ** (_ ** (b1, b1, _)))
\end{verbatim}
\end{center}
\caption{Addition of three bits}
\figrule
\end{figure}

@@

Now, we want to define a function adding two binary numbers and a bit (used as a carry). We decide that this addition will only work for two binary numbers expressed with the same number of bits, and we produce a result with one more bit. This result should of course represent the value $c + x + y$ if $c$ is representing the value of the input bit, and if $x$ and $y$ are representing the values of the two binary numbers given in input. Thus, we're tempted to write :
\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
adc : Binary w x -> Binary w y -> Bit c -> Binary (S w) (c + x + y)
adc zero zero carry = zero # carry
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = addBit bX bY carry in
          adc numx numy carry0 # lsb
\end{verbatim}
\end{center}
\caption{Addition of two binary numbers}
\figrule
\end{figure}

Unfortunately, the types are mismatching for the two cases expressed by this pattern matching. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type\[Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)\] but we're providing a term of type \[Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))\] [I AM SURPRISED HERE BY THE TYPE OF THE PROVIDED TERM].
The problem is even worse for the second line where the result is expected to have type \[plus\ (plus\ c\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0)))\] while we're trying to provide a term of type \[plus\ v2\ (plus\ (plus\ (plus\ v1\ v)\ v)\ (plus\ (plus (plus\ v1\ v)\ v)\ 0))\]
For this kind of situation, Idris provides the possibility to write a provisional definition with the syntax "?=". Using a provisional definitions here will make the definition being accepted, but it also generates two proof obligations, one for each pattern, and both of them are saying that the provided term can be converted into a term which has the expected type.
For both of these two patterns, it is in fact possible to prove the equality between the expected and the provided types, because this function is correct. These proofs are made of a series of rewriting, and each rewriting uses a property about the operation $+$ on natural numbers.
For example, the proof for the second pattern looks like this :
\begin{figure}[H]
\figrule
\begin{center}
\begin{verbatim}
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
\end{verbatim}
\end{center}
\caption{Proof by hand of the correspondence between expected and provided types for the second pattern of the definition of adc}
\figrule
\end{figure}

@@

Algebraic provers, like ring provers, are already implemented for various proof assistants, including Coq and Agda. In this paper, we describe a certified implementation of an automatic prover for equalities in various algebraic structures, including Monoid, Groups and Rings (all potentially commutative), for the Idris language. The novelty is in the approach that we follow, using a new kind of type-safe reflection.
Working by reflection for implementing tactics has been done several times, as in [ref 1] for the implementation of a ring solver for the Coq proof assistant, but their reflection mechanism  doesn't provide any guarantee, contrary to our type-safe reflection. We will compare our approach with other implementation of this kind of tactics in section 5. \\
\\
Our contribution mainly consists of three points :
\begin{itemize}
	\item 1) We present a new type-safe reflection mechanism, where the reflected terms are indexed over the concrete inputs, thus providing an easy way to pull out the proofs easily, and providing two guarantees. The first one is that the reflection of a term is guaranteed to be a faithful representation of the term. The second is that any generated proof is guaranteed to be a proof of the required property. The basic ideas of the technique are first presented in the section 2 on a smaller problem with only natural numbers and additions, and are later adapted for the hierarchy of tactics in section 3 (and section 4 contains some implementation details).	
	\item 2) The normalisation procedures are implemented by following a correct by construction approach, instead of implementing a normalization procedure, and proving afterwards that this function effectively computes a normal form. This approach is much more suitable for programming languages like Idris. Again, this approach will be presented in section 2 on the smaller problem, and on sections 3 and 4 for the complete hierarchy of algebraic structures.
	\item 3) We develop a hierarchy of tactics which enables to increase a lot the reusability of each prover : we reuse as much as possible each prover in the implementation of the other solvers for the above structures. For example simplifying neutral elements is only implemented at the monoid level, and each level above monoid reuses it. As it will be discussed in section 4, re-usability can be quite tricky to get when we want to reuse a solver from a structure above which is less expressive. For example, reusing the monoid solver for building the group solver is not trivial, since we lose the possibility to express negations ($-x$) and minus operations ($x-y$).
\end{itemize}

>>>>

For example, when working with natural numbers, one might need to prove the lemma $\forall\ xy:Nat,\ x*x\ +\ 3*x\ +\ x*y\ =\ x*(3+y+x)$, and when working with lists and concatenation of lists, one might need to prove something like $\forall\ l1\ l2\ l3:List A,\ ((l1\ ++\ (l2 ++ l3))\ ++\ l4\ =\ (l1++l2)\ ++\ (l3++l4)$. Proving this kind of auxiliary lemmas is time consuming, is much too specific for being reusable, and isn't the interesting part of formal certification. Therefore, we would like to be able to automatize a far as possible the constructions of this kind of proofs for the Idris language, a dependently typed language which doesn't contain many proof automations at the moment. \\
\\
When doing formal certification, many functions have a lemma of correctness that talks about equality of terms. More precisely, the following situation is very familiar. The function that we wants to certify is $f\ :\ A \rightarrow B$. We have two functions $sem_A\ :\ A \rightarrow SEM$ and $sem_B\ : \ B \rightarrow SEM$ of interpretation which associate a semantic to each type A and B. If this semantic says everything that matters for the behaviour of $f$, then the lemma of correctness of f can be expressed as $forall\ a:A,\ semA\ a = semB (f\ a)$. The interesting thing is that, very often, the type $SEM$ has the structure of some nice algebraic structure, and it becomes thus possible to decide equalities in it automatically.
All the equality symbols we have encountered so far refer to the propositional equality. We remind the readers in the next subsection how this equality is defined in Idris, and more generally in Intentional Type Theories, and how it differs from the --more primitive-- judgemental equality.

@@

The judgemental equality has to be included in the propositional equal, because what is equal by definition should be provably equal.
This is accomplished by giving constructors to the type $Id(a,a)$ and nothing when "$a$ is not $b$".
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
Id (A:Type) : A -> A -> Type :=
  Refl : forall a:A, Id a a
\end{lstlisting}
\end{center}
\caption{The type Id}
\figrule
\end{figure}

@@

\subsection{A motivating little example}
We introduce a little example which shows how these proof obligations appear when we index a type by values like natural numbers.
Let's define the types of $Bit$ and $Binary$ number, both indexed over the value they represent, expressed in base ten as a natural number.
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
data Bit : Nat -> Type where
     b0 : Bit Z
     b1 : Bit (S Z)

data Binary : (width : Nat) -> (value : Nat) -> Type where
     zero : Binary Z Z
     (#) : Binary w v -> Bit bit -> Binary (S w) (bit + 2 * v)
\end{lstlisting}
\end{center}
\caption{Bit and binary number}
\figrule
\end{figure}

@@

\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
addBit : Bit x -> Bit y -> Bit c -> (bX ** (bY ** 
              (Bit bX, Bit bY, c + x + y = bY + 2 * bX)))
addBit b0 b0 b0 = (_ ** (_ ** (b0, b0, _)))
addBit b0 b0 b1 = (_ ** (_ ** (b0, b1, _)))
addBit b0 b1 b0 = (_ ** (_ ** (b0, b1, _)))
addBit b0 b1 b1 = (_ ** (_ ** (b1, b0, _)))
addBit b1 b0 b0 = (_ ** (_ ** (b0, b1, _)))
addBit b1 b0 b1 = (_ ** (_ ** (b1, b0, _)))
addBit b1 b1 b0 = (_ ** (_ ** (b1, b0, _)))
addBit b1 b1 b1 = (_ ** (_ ** (b1, b1, _)))
\end{lstlisting}
\end{center}
\caption{Addition of three bits}
\figrule
\end{figure}

@@

Now, we want to define a function adding two binary numbers and a bit (used as a carry). We decide that this addition will only work for two binary numbers expressed with the same number of bits, and we produce a result with one more bit. This result should of course represent the value $c + x + y$ if $c$ is representing the value of the input bit, and if $x$ and $y$ are representing the values of the two binary numbers given in input. Thus, we're tempted to write :
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
adc : Binary w x -> Binary w y -> Bit c -> Binary (S w) (c + x + y)
adc zero zero carry = zero # carry
adc (numx # bX) (numy # bY) carry
   = let (v1 ** (v2 ** (carry0, lsb, _))) = addBit bX bY carry in
          adc numx numy carry0 # lsb
\end{lstlisting}
\end{center}
\caption{Addition of two binary numbers and a carry}
\figrule
\end{figure}

Unfortunately, the types are mismatching for the two cases expressed by this pattern matching. The result of the first line $adc\ zero\ zero\ carry$ is expected to have the type\[Binary\ 1\ (plus\ (plus\ c\ 0)\ 0)\] but we're providing a term of type \[Binary\ 1\ (plus\ c\ (plus\ 0\ (plus\ 0\ 0)))\] [I AM SURPRISED HERE BY THE TYPE OF THE PROVIDED TERM].
The problem is even worse for the second line where the result is expected to have type \[plus\ (plus\ c\ (plus\ bit\ (plus\ v\ (plus\ v\ 0))))\ (plus\ bit\ (plus\ v\ (plus\ v\ 0)))\] while we're trying to provide a term of type \[plus\ v2\ (plus\ (plus\ (plus\ v1\ v)\ v)\ (plus\ (plus (plus\ v1\ v)\ v)\ 0))\]
For this kind of situation, Idris provides the possibility to write a provisional definition with the syntax "?=". Using a provisional definitions here will make the definition being accepted, but it also generates two proof obligations, one for each pattern, and both of them are saying that the provided term can be converted into a term which has the expected type.
For both of these two patterns, it is in fact possible to prove the equality between the expected and the provided types, because this function is correct. These proofs are made of a series of rewriting, and each rewriting uses a property about the operation $+$ on natural numbers.
For example, the proof for the second pattern looks like this :
\begin{figure}[H]
\figrule
\begin{center}
\begin{lstlisting}
adc_lemma_2 = proof {
    intros;
    rewrite sym (plusZeroRightNeutral x);
    [...]
    rewrite sym (plusAssociative (plus c (plus bit0 (plus v v))) bit1 (plus v1 v1));
    rewrite (plusAssociative c (plus bit0 (plus v v)) bit1);
    rewrite plusCommutative bit1 (plus v v);
    [...]
    rewrite (plusAssociative (plus (plus x v) v1) (plus x v) v1);
    trivial;
\end{lstlisting}
\end{center}
\caption{Proof by hand of the correspondence between expected and provided types for the second pattern of the definition of adc}
\figrule
\end{figure}

@@

Algebraic provers, like ring provers, are already implemented for various proof assistants, including Coq and Agda. In this paper, we describe a certified implementation of an automatic prover for equalities in various algebraic structures, including Monoid, Groups and Rings (all potentially commutative), for the Idris language. The novelty is in the approach that we follow, using a new kind of type-safe reflection.
Working by reflection for implementing tactics has been done several times, as in [ref 1] for the implementation of a ring solver for the Coq proof assistant, but their reflection mechanism  doesn't provide any guarantee, contrary to our type-safe reflection. We will compare our approach with other implementations of this kind of tactics in section 5. \\
\\
Our contribution mainly consists of three points :
\begin{itemize}
	\item 1) We present a new type-safe reflection mechanism, where the reflected terms are indexed over the concrete inputs, thus providing an easy way to pull out the proofs easily, and providing two guarantees. The first one is that the reflection of a term is guaranteed to be a faithful representation of the term. The second is that any generated proof is guaranteed to be a proof of the required property. The basic ideas of the technique are first presented in the section 2 on a smaller problem with only natural numbers and additions, and are later adapted for the hierarchy of tactics in section 3 (and section 4 contains some implementation details).	
	\item 2) The normalisation procedures are implemented by following a correct by construction approach, instead of implementing a normalization procedure, and proving afterwards that this function effectively computes a normal form. This approach is much more suitable for programming languages like Idris. Again, this approach will be presented in section 2 on the smaller problem, and on sections 3 and 4 for the complete hierarchy of algebraic structures.
	\item 3) We develop a hierarchy of tactics which enables to increase a lot the reusability of each prover : we reuse as much as possible each prover in the implementation of the other solvers for the above structures. For example simplifying neutral elements is only implemented at the monoid level, and each level above monoid reuses it. As it will be discussed in section 4, re-usability can be quite tricky to get when we want to reuse a solver from a structure above which is less expressive. For example, reusing the monoid solver for building the group solver is not trivial, since we lose the possibility to express negations ($-x$) and minus operations ($x-y$).
\end{itemize}
